\CRANpkg{lg}: An R package for Local Gaussian Approximations

### 2021-02-16 response/response.Rmd

> *To exemplify this, the section on ‘Statistical background’ reads as if compiled from different sources; In particular, the paragraphs are not well connected in my view.*

**My response:** The purpose of this section is to give the reader a brief overview of the statistical methods that have been implemented in the **lg**-package, without going too much in to the technical details. I have revised this section carefully, with particular emphasis on improving the transitions between the paragraphs.

>  *Further the explanation of the basic idea of the GLC measure in the section ‘Statistical background’ could be improved, thereby making the concept accessible for a broader audience and making this paper more ‘self-contained’.*

**My response:** Thank you for this input. The reviewer is correct to point out that the section on "Statistical background" is fairly short and does not go into great detail. The composition of this section is the result of a careful trade-off. On the one hand I fully agree that a more comprehensive section would be desirable, because that would allow the reader to appreciate the technical nuances within the topic of the LGC in greater detail. On the other hand, it is the purpose of this paper to demonstrate the practical implementation of a large and diverse set of statistical methods. It is my conclusion that this section should not be much longer than it is in its current form (approx. 4 pages), which naturally puts a limit on the level of detail. I have instead opted to provide a rather accessible birds-eye view of the relevant literature, which I believe is useful to the readership of *The R Journal*. This section has been heavily revised as prompted by the reviewer in his/her comment above.

@@

> *Also, the whole manuscript is not well-structured: The section with the illustration of the lg package is called ‘The lg-object’ and more importantly is confusing since, for instance, all three different estimation methods implemented are outlined, although not necessary for the (empirical) example chosen by the author;*

**My reponse:** The paper is structured in the same way as the package it describes. There are two steps to every analysis; in the first step the practitioner must make several modeling choices which to a large degree is guided by the respective references. This step is common to all applications. I have created a new title for this section to clarify the structure. It is now called "The first step: Creating the lg-object". The various applications are then described separately and organized under informative headlines.

@@

I am not entirely sure how to address the reviewer's second remark. The examples in this paper are chosen to illustrate key steps in the implementation of the methods. It is not my intention to demonstrate all possible combinations of the modeling choices in examples, as that would likely reduce the value of the paper as a general, readable and fairly brief introduction to the relevant R-functions.

> *A section with a summary or concluding remarks is fully missing!*

**My response:** I have added some concluding remarks in the end of the paper.

@@

> *1. As far as I see is your manuscript missing a final section that summarizes the main ideas (e.g. functions here) from the paper and provides some concluding remarks and perhaps an outlook (e.g. what will be added to package in the future). I strongly recommend to provide such a section.*

**My response:** Has been added.

> *2. Are you sure that Equation (1) is correct? I think $x$ should be $x_0$ . Further I do not understand $y$ and assume $y$ should be $X_i$?*

**My response:** Yes, this equation is correct. The $x$ is the location of the local estimate as defined in the discussion leading up to equation (1), and $y$ is the integration variable in the penalty term of the local likelihood function. This notation is consistent with the literature on local likelihood estimation. Indeed, the second term does not depend on the data, and the reviewer is quite right to point out that this is not obvious. I have therefore added a comment about this, as well as a reference to a much more detailed justification of the local likelihood function.

@@

> *3. Are you sure that ‘the same test statistic and bootstrap procedure as for independence’ can be used with time series data? Is not block-bootstrap necessary in this case? Please explain briefly! (p. 3)*

**My response:** In this particular case we can use the standard bootstrap, because serial dependence in a time series can be investigated by considering the statistical dependence between $X_t$ and $X_{t-k}$ directly. This is a parallel to the autocorrelation function, which for a stationary time series is defined as the ordinary Pearson correlation $\gamma_k = Corr(X_t, X_{t-k})$. I have changed the formulation here so that it directly states that we obtain critical values by permutations. The block bootstrap is then introduced in the next paragraph, where it is needed to test for serial cross-dependence.

@@

> *4. I understand that the bandwidth controls how many observations are used to compute the correlation coefficient of one grid. Does lg also allow to extract the (exact) number of observations used to compute the different correlations coefficients (as depicted e.g. in Fig. 1)? I feel that from the perspective of an applied statistician, this figure could provide an intuition about how reliable one or the other coefficient is.*

**My response:** The bandwidth does not define a finite region over which the usual correlation is calculated. The LGC is not the conditional correlation, and the bandwidth does not correspond to an exact number of observations. Figure 1 shows the estimated local Gaussian correlation plotted out on a grid chosen by the user, and I have pointed this out in the first line of the paragraph that introduces this figure.

Although one can discuss the "effective number of observations" in the classical nonparametric sense, this has not been done in the literature on the LGC, and hence is not a concept in the *lg*-package.

@@

**My response:** That may certainly be possible. To the best of my knowledge, however, this is not a test that has been introduced in the literature. It is therefore not implemented in the **lg**-package.

> *6. Page 1: I suggest to change the statement ‘clever and creative techniques’ by a more scientific formulation.*

**My response:** This statement has been taken out. Thank you.

I addition to the changes prompted by the reviewer, I have updated a few references and made several minor adjustments to the text throughout the article to improve the presentation.

### 2021-02-16 otneim.tex
Full pass with spelling of the entire paper and response to reviewer

\abstract{
The package \CRANpkg{lg} for the R programming language provides implementations of recent methodological advances on applications of the local Gaussian correlation. This includes estimation of the local Gaussian correlation itself, multivariate density estimation, conditional density estimation, various tests for independence and conditional independence, as well as a graphical module for creating dependence maps. This paper describes the \CRANpkg{lg} package, its principles and its practical use.
}

\section{Introduction}

\citet{tjos:huft:2013} propose the \emph{local Gaussian correlation} (LGC) as a new measure of statistical dependence between two stochastic variables $X_1$ and $X_2$, which has the following important property yet unrivaled in the literature: It can separate between positive and negative nonlinear dependence while still reducing to the ordinary Pearson's correlation coefficient if $X_1$ and $X_2$ are jointly normally distributed. The R-package \CRANpkg{localgauss} \citep{bere:klep:tjos:2014} provides two important functions in this context; one that calculates the sample LGC based on observed values of $(X_1,X_2)$, and one that uses the estimated LGC to perform a local test of independence between $X_1$ and $X_2$ as described in detail by \citet{bere:tjos:2014}.

@@

This paper describes the \CRANpkg{lg}-package \citep{otne:2019}, which provides a unified framework to implement all these methods, as well as a tool for visualizing the LGC and LGPC as dependence maps (and conditional dependence maps). \citet{jord:tjos:2017} use the LGC in spectral analysis of time series, but those methods have their own computational ecosystem in the \pkg{localgaussSpec}-package, \citet{jord:2018}.

@@

A different approach is to estimate $f_X(\cdot)$ without any prior parametric assumptions. The classical method for non-parametric denstity estimation is the kernel estimator 
$$\widehat f_X(x) = \frac{1}{nb}\sum_{i=1}^n K\left(\frac{X_i - x}{b}\right),$$
where $K$ is a symmetric density function (the kernel) and $b$ is a tuning parameter (the bandwidth) that controls the smoothness of the estimate $\widehat f_X(\cdot)$. See \citet{silv:1986} for an introduction to this topic. There is also a massive statistical literature on density estimation containing extensions and improvements to the classical methods to be used in various practical situations.

\citet{hjor:jone:1996} provide one such idea. They consider a parametric family $F_{\theta}$, but instead of searching for a single parameter value $\theta_0$, for which $f_X(x) = f(x;\theta_0)$ (or approximately so), they rather assert that different members of $F_{\theta}$ may approximate $f_X$ locally in different parts of its domain. In other words, they seek to estimate a parameter \emph{function} $\theta_0(x)$ for which $f_X(x) = f(x;\theta_0(x))$ (or approximately so), and do this by maximizing a \emph{local likelihood function} in each point $x$:
\begin{align}
\widehat\theta(x) 
&= \arg\max_{\theta \in \Theta} L_n(\theta, x) \nonumber\\ 
&= \arg\max_{\theta \in \Theta} \frac{1}{nb}\sum_{i=1}^n K\left(\frac{X_i - x}{b}\right)\log f(X_i; \theta) - \int \frac{1}{b}K\left(\frac{y - x}{b}\right)f(y;\theta)\,\textrm{d}y,
\label{eq:loclik}
\end{align}
where, again, $K$ is a symmetric density function and $b$ is a bandwidth parameter that controls the smoothness of the estimate. The second term in the local likelihood function is a penalty that ensures that the estimated density $\widehat f_X(x) = f(x; \widehat\theta(x))$ converges correctly to the true density function $f_X(x)$ as the sample size $n$ increases to infinity and the bandwidth $b$ decreases towards zero. See \citet{hjor:jone:1996} for a detailed discussion about this construction.

@@

In Figure \ref{fig:initial-example} we have plotted the estimated LGC for two bivariate data sets on a grid; 1000 simulated observations from a binormal distribution having correlation equal to 0.5, and the daily return on the CAC40 and FTSE100 stock indices on 1000 consecutive trading days starting on May 5th 2014. In the first panel, we see that the estimated local correlation coincides with the global correlation, except for the estimation error. In the second panel we see clearly that the local correlation, and thus the dependence, is stronger in the lower left and upper right regions of the distribution than in central parts. The phenomenon of local dependence is well known in the financial literature, and using the LGC it can be measured, interpreted, and visualized in a natural way. The interpretation of this particular figure is that extreme observations on the two stock indices are more strongly dependent than the less extreme observations.

@@

Consider next the stationary time series $\{X_t\}$. The autocorrelation function (ACF) $\rho_k = \rho(X_t, X_{t-k})$ is a well known concept for describing the serial dependence in the time seris, but the ACF is, again, only capable to completely capture \emph{linear} serial dependence. \citet{laca:tjos:2017} seek to remedy this by rather calculating the the local correlation between $X_t$ and $X_{t-k}$. This leads to a test for serial independence in the natural way. In fact, this work is mainly a theoretical exercise in order to accommodate time series dependence. Testing for independence between $X_t$ and $X_{t-k}$ using observations $\{X_t, X_{t-k}\}_{t = k+1}^T$ leads to the same test statistic \eqref{eq:test-statistic1} and bootstrap procedure as the test for independence between $X_1$ and $X_2$ that we described above.

@@

Although the original work by \citet{hjor:jone:1996} provide a general framework for local likelihood density estimation, using any $p$-variate parametric family as the local family, it is evident that the method may struggle in multivariate applications much in the same way as the kernel density estimator does. This is a consequence of the curse of dimensionality, the effect of which is sought remedied by an algorithm provided by \citet{otne:tjos:2017}. The idea is to fit the $p$-variate normal distribution $\psi(\mu, \Sigma)$ locally, where $\mu$ is the vector of $p$ expectations, and $\Sigma$ is the $p\times p$ covariance matrix (to which the correlation matrix $R$ corresponds), but under the following structural simplifications:

\begin{align}
\mu_i(x) &= \mu_i(x_1, \ldots, x_p) \stackrel{\textrm{def}}{=} \mu_(x_i) \label{eq:simp-mu}\\
\sigma_i(x) &= \sigma_i(x_1, \ldots, x_p)  \stackrel{\textrm{def}}{=} \sigma_i(x_i) \label{eq:simp-sig} \\
\rho_{ij}(x) &= \rho_{ij}(x_1, \ldots, x_p) \stackrel{\textrm{def}}{=} \rho_{ij}(x_i, x_j). \label{eq:simp-rho}
\end{align}
\citet{otne:tjos:2017} estimate the local parameters above by first obtaining univariate marginal locally Gaussian fits (eqs. \ref{eq:simp-mu} and \ref{eq:simp-sig}), and then pairwise bivariate locally Gaussian fits (eq. \ref{eq:simp-rho}). In the second step the estimates $\widehat\mu_i(x_i)$, $\widehat\mu_j(x_j)$, $\widehat\sigma_i(x_i)$ and $\widehat\sigma_j(x_j)$ are kept fixed in the estimation of the pairwise local correlation. They argue further that the following transformation technique will produce better density estimates in many situations. Denote by $F_i(x_i)$, $i = 1,\ldots, p$ the marginal distribution functions of the stochastic vector $X$, and by $\widehat F_i(x_i) = n^{-1}\sum_{i=1}^n 1(X_i \leq x_i)$ their empirical counterparts. They then estimate the density $f_Z(z)$ of the vector $Z = \{\Phi^{-1}(F_i(X_i))\}_{i=1,\ldots,p}$, in practice approximated by

@@

\begin{equation}
\mu_i(z_i) \stackrel{\textrm{def}}{=} 0 \,\, \textrm{ and } \sigma_i(z_i) \stackrel{\textrm{def}}{=} 1, \,\, i = 1,\ldots,p,
\label{eq:simp}
\end{equation}
so that the only parameter functions left to estimate are the pairwise local Gaussian correlations $R(z) = \{\rho_{ij}(z_i, z_j)\}_{i<j}$. We use the notation $Z$, $z_i$ and $z_j$ to signify that the estimation is performed on the (approximate) standard normal scale, or $z$-scale for short. We can then estimate the joint density $f_Z(z)$ of $Z$ as
\begin{equation}
\widehat f_Z(z) = \psi(z; \mu = 0, \sigma = 1, R = \widehat R(z)),
\label{eq:transformed-density}
\end{equation}
and then substitute $f_Z$ for $\widehat f_Z$ in the following relation obtained by \citet{otne:tjos:2017} in order to estimate the unknown density $f_X$:
\begin{equation}
f(x) = f_{Z}\big(\Phi^{-1}(F_1(x_1)), \ldots, \Phi^{-1}(F_p(x_p))\big) \times \prod_{i=1}^p \frac{f_i(x_i)}{\phi\big(\Phi^{-1}(F_i(x_i))\big)},
\label{eq:backtrans}
\end{equation}
where $\phi(\cdot)$ is the standard normal pdf. This estimator is implemented the \CRANpkg{lg}-package as demonstrated in Section \ref{chap:density}.

@@

\begin{align}
\mu^*(z) &= R_{12}(z)R_{22}(z)^{-1}z^{(2)},  \label{eq:cond-mu-trans}\\
\Sigma^*(z) &= R_{11}(z) - R_{12}(z)R_{22}(z)^{-1}R_{21}(z), \label{eq:cond-sig-trans}
\end{align}
where we, again, switch to $z$-notation in order to make it clear that these quantities are estimated on the standard normal $z$-scale. An estimator $\widehat f_{X^{(1)}|X^{(2)}}(\cdot|\cdot)$ of the conditional density $f_{X^{(1)}|X^{(2)}}(\cdot|\cdot)$ follows immediately from an expression corresponding to (\ref{eq:backtrans}), and the \CRANpkg{lg}-package provides functions for implementing this estimator in R, and we describe the implementation of this functionality in Section \ref{chap:density}.

Finally, we refer to \citet{otne:tjos:2019}, who take the local version of the conditional covariance matrix (\ref{eq:cond-sig}) (or (\ref{eq:cond-sig-trans}) in the transformed case) as a measure for conditional dependence, and thus as an instrument to test for conditional independence. Consider the stochastic vector $X = \left(X^{(1)}, X^{(2)}, X^{(3)}\right)$, where $X^{(1)}$ and $X^{(2)}$ are scalars and $X^{(3)}$ may be a vector. $X^{(1)}$ is conditionally independent from $X^{(2)}$ given $X^{(3)}$, written $X^{(1)} \perp X^{(2)} \,|\,X^{(3)}$, if the stochastic variables $X^{(1)}\,|\,X^{(3)}$ and $X^{(2)}\,|\,X^{(3)}$ are independent, or, equivalently, if the joint conditional density of $X^{(1)}$ and $X^{(2)}$ given $X^{(3)}$ can be written as the product

@@

The local Gaussian correlation may be used to perform a number of statistical analyses, as is evident from the preceding section. The practitioner must first, however, make three quite specific modelling choices; namely (i) to choose an estimation method, i.e. the level of simplification in multivariate applications, (ii) to determine whether the data should be transformed towards marginal standard normality before estimating the LGC, and (iii) to choose a set of bandwidths, or at least a method for calculating bandwidths. The architecture of the \pkg{lg}-packages requires the user to make these choices \emph{before} endeavoring further into specific applications by imposing a strict, two step procedure:

@@

In the following we assume that one has a \emph{data set} \code{x} loaded into the R workspace, which must be an $n\times p$ matrix (one column per variable, one row per observation), possibly including \code{NA}`s which will be excluded, or a data frame having the same dimensions. The fundamental syntax for creating an lg-object is \code{lg\_object <- lg\_main(x)}, and we will in this section explain how the modelling decisions (i)-(iii) can be encoded into the lg-object.

@@

Note that the full locally Gaussian fit for raw data is not available if the number of variables $p$ is greater than 2. The \code{lg\_main()}-function will check for this and throw an error if $p>2$ and \code{est\_method = "5par"}.

@@

\textbf{3. A simplified locally Gaussian fit for marginally standard normal data.} This estimation method is applicable for marginally standard normal data, or data that have been transformed to approximate marginal standard normality by e.g. the transformation (\ref{eq:trans}). In that case, we fix the marginal expectation functions and standard deviation functions to the constant values $0$ and $1$ respectively, rather then estimating them. To use this estimation method, create the lg-object by running

@@

Note that the function call above will issue a warning if the option for transforming the data to marginal standard normality is not at the same time set to \code{TRUE}, see the next sub-section for details.

\textbf{4. A full locally Gaussian fit for trivariate data.} If $p = 3$ \emph{and} we choose to transform the data to marginal standard normality (see the next sub-section), the transformed density $f_Z(\cdot)$ in (\ref{eq:transformed-density}) may be estimated by jointly estimating the three local correlations $\rho_{12}(z_1, z_2, z_3)$, $\rho_{13}(z_1, z_2, z_3)$ and $\rho_{23}(z_1, z_2, z_3)$. This estimation method was introduced recently by \citet{otne:tjos:2019} in order to increase power of their conditional independence test, but it can be used in any application described in this paper that consider trivariate data. To use this estimation method, create the lg-object by running

@@

We have in practice seen two automatic bandwidth selectors employed in the applications referred to in Section \ref{chap:background}: a
cross-validation procedure, which is slow, but accurate with respect to density estimation, and a plug-in bandwidth that is quick, but cruder. We use the argument
\code{bw\_method} to the \code{lg\_main()}-function in order to choose between the two.

@@

$joint
  x1 x2       bw1      bw2 convergence
1  1  2 0.2946889 0.331971           0
\end{example}
This is itself a list, containing the crucial elements \code{marginal} for the \(p\) marginal bandwidths, and \code{joint} that contains the
\(p(p-1)/2\) bandwidth matrices, one for each pair of variables (which in this bivariate example just \emph{one} variable pair, (\code{x1,\ x2})). The convergence flags stem from the \code{optim()} and \code{optimize()} functions that we use to obtain the minimizer of \(CV(\cdot)\), and 0 indicates successful convergence.

@@

In the sub-section above we present the three most important arguments to \code{lg\_main()}. Each of them allows the user to configure one of the three crucial modelling choices. Let us complete this treatment by covering some possibilities to make further adjustments to those choices.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The user may supply the bandwidths directly to \code{lg\_main()} by passing them to the \code{bw}-argument. They have to be in the   correct format though, which is a list containing the vector \code{\$marginal} if \code{est\_method\ =\ 5par\_marginals\_fixed}, and always a data frame \code{\$joint} specifying all variable pairs in the \code{x1} and \code{x2} columns and the corresponding bandwidths
  in the \code{bw1} and \code{bw2} columns. The function \code{bw\_simple()} will assist in creating bandwidth objects.
\item
  If \code{bw\_method\ =\ "plugin"} the user may change the proportionality constant and exponent in the plugin formula for the joint and, if applicable, the marginal bandwidths. See Table \ref{tab:arguments-lg-main} for the necessary argument names.
\item
  If \code{bw\_method\ =\ "cv"}, the user may change the numerical tolerance in the optimization of \(CV(b)\). See Table \ref{tab:arguments-lg-main} for the necessary argument names.
\end{enumerate}

@@

We proceed in this section to demonstrate how to implement each of the tasks that we discussed in Section \ref{chap:background}. The general pattern is to pass an lg-object, see the preceding section, to one of the estimation or test functions provided in the \CRANpkg{lg}-package. We will look at some financial data in the examples: the monthly returns on the S\&P500, FTSE100, DAX30 and TOPIX stock indices from January 1985 to March 2018. This data is available as an online appendix to this article.

>>>>

\abstract{
The package \CRANpkg{lg} for the R programming language provides implementations of recent methodological advances on applications of the local Gaussian correlation. This includes estimation of the local Gaussian correlation itself, multivariate density estimation, conditional density estimation, various tests for independence and conditional independence, as well as a graphical module for creating dependence maps. This paper describes the \CRANpkg{lg} package, its principles, and its practical use.
}

\section{Introduction}
\label{chap:introduction}

\citet{tjos:huft:2013} propose the \emph{local Gaussian correlation} (LGC) as a new measure of statistical dependence between two stochastic variables $X_1$ and $X_2$, which has the following important property yet unrivaled in the literature: It can separate between positive and negative nonlinear dependence while still reducing to the ordinary Pearson correlation coefficient if $X_1$ and $X_2$ are jointly normally distributed. The R-package \CRANpkg{localgauss} \citep{bere:klep:tjos:2014} provides two important functions in this context; one that calculates the sample LGC based on observed values of $(X_1,X_2)$, and one that uses the estimated LGC to perform a local test of independence between $X_1$ and $X_2$ as described in detail by \citet{bere:tjos:2014}.

@@

This paper describes the \CRANpkg{lg}-package \citep{otne:2019}, which provides a unified framework to implement all these methods, as well as a tool for visualizing the LGC and LGPC as dependence maps. \citet{jord:tjos:2017} use the LGC in spectral analysis of time series, but those methods have their own computational ecosystem in the \pkg{localgaussSpec}-package, \citet{jord:2018}.

@@

A different approach is to estimate $f_X(\cdot)$ without any prior parametric assumptions. The classical method for non-parametric density estimation is the kernel estimator 
$$\widehat f_X(x) = \frac{1}{nb}\sum_{i=1}^n K\left(\frac{X_i - x}{b}\right),$$
where $K$ is a symmetric density function (the kernel) and $b$ is a tuning parameter (the bandwidth) that controls the smoothness of the estimate $\widehat f_X(\cdot)$. See \citet{silv:1986} for an introduction to this topic. There is also a massive statistical literature on density estimation containing extensions and improvements to the classical methods to be used in various practical situations.

\citet{hjor:jone:1996} provide one such idea. They consider a parametric family $F_{\theta}$, but instead of searching for a single parameter value $\theta_0$, for which $f_X(x) = f(x;\theta_0)$ (or approximately so), they rather assert that different members of $F_{\theta}$ may approximate $f_X$ locally in different parts of its domain. In other words, they seek to estimate a parameter \emph{function} $\theta_0(x)$ for which $f_X(x) = f(x;\theta_0(x))$ (or approximately so), and do this by maximizing a \emph{local likelihood function} in each point $x$:
\begin{align}
\widehat\theta(x) 
&= \arg\max_{\theta \in \Theta} L_n(\theta, x) \nonumber\\ 
&= \arg\max_{\theta \in \Theta} \frac{1}{nb}\sum_{i=1}^n K\left(\frac{X_i - x}{b}\right)\log f(X_i; \theta) - \int \frac{1}{b}K\left(\frac{y - x}{b}\right)f(y;\theta)\,\textrm{d}y,
\label{eq:loclik}
\end{align}
where, again, $K$ is a symmetric density function and $b$ is a bandwidth parameter that controls the smoothness of the estimate. The second term in the local likelihood function is a penalty that ensures that the estimated density $\widehat f_X(x) = f(x; \widehat\theta(x))$ converges correctly to the true density function $f_X(x)$ as the sample size $n$ increases to infinity, and the bandwidth $b$ decreases towards zero. See \citet{hjor:jone:1996} for a detailed discussion about this construction.

@@

In Figure \ref{fig:initial-example} we have plotted the estimated LGC for two bivariate data sets on a grid; 1000 simulated observations from a binormal distribution having correlation equal to 0.5, and the daily return on the CAC40 and FTSE100 stock indices on 1000 consecutive trading days starting on May 5th 2014. In the first panel, we see that the estimated local correlation coincides with the global correlation, except for the estimation error. In the second panel we see clearly that the local correlation, and thus the dependence, is stronger in the lower left and upper right regions of the distribution than in the central parts. The phenomenon of local dependence is well known in the financial literature, and using the LGC it can be measured, interpreted, and visualized in a natural way. The interpretation of this particular figure is that extreme observations on the two stock indices are more strongly dependent than the less extreme observations.

@@

Consider next the stationary time series $\{X_t\}$. The autocorrelation function (ACF) $\rho_k = \rho(X_t, X_{t-k})$ is a well known concept for describing the serial dependence in the time series, but the ACF is, again, only capable to completely capture \emph{linear} serial dependence. \citet{laca:tjos:2017} seek to remedy this by rather calculating the the local correlation between $X_t$ and $X_{t-k}$. This leads to a test for serial independence in the natural way. In fact, this work is mainly a theoretical exercise in order to accommodate time series dependence. Testing for independence between $X_t$ and $X_{t-k}$ using observations $\{X_t, X_{t-k}\}_{t = k+1}^T$ leads to the same test statistic \eqref{eq:test-statistic1} and bootstrap procedure as the test for independence between $X_1$ and $X_2$ that we described above.

@@

Although the original work by \citet{hjor:jone:1996} provide a general framework for local likelihood density estimation using any $p$-variate parametric family as the local family, it is evident that the method may struggle in multivariate applications much in the same way as the kernel density estimator does. This is a consequence of the curse of dimensionality, the effect of which is sought remedied by an algorithm provided by \citet{otne:tjos:2017}. The idea is to fit the $p$-variate normal distribution $\psi(\mu, \Sigma)$ locally, where $\mu$ is the vector of $p$ expectations, and $\Sigma$ is the $p\times p$ covariance matrix (to which the correlation matrix $R$ corresponds), but under the following structural simplifications:

\begin{align}
\mu_i(x) &= \mu_i(x_1, \ldots, x_p) \stackrel{\textrm{def}}{=} \mu_(x_i) \label{eq:simp-mu}\\
\sigma_i(x) &= \sigma_i(x_1, \ldots, x_p)  \stackrel{\textrm{def}}{=} \sigma_i(x_i) \label{eq:simp-sig} \\
\rho_{ij}(x) &= \rho_{ij}(x_1, \ldots, x_p) \stackrel{\textrm{def}}{=} \rho_{ij}(x_i, x_j). \label{eq:simp-rho}
\end{align}
\citet{otne:tjos:2017} estimate the local parameters above by first obtaining univariate marginal locally Gaussian fits (eqs. \ref{eq:simp-mu} and \ref{eq:simp-sig}), and then pairwise bivariate locally Gaussian fits (eq. \ref{eq:simp-rho}). In the second step the estimates $\widehat\mu_i(x_i)$, $\widehat\mu_j(x_j)$, $\widehat\sigma_i(x_i)$ and $\widehat\sigma_j(x_j)$ are kept fixed in the estimation of the pairwise local correlation. They argue further that the following transformation technique will produce better density estimates in many situations.

Denote by $F_i(x_i)$, $i = 1,\ldots, p$ the marginal distribution functions of the stochastic vector $X$, and by $\widehat F_i(x_i) = n^{-1}\sum_{i=1}^n 1(X_i \leq x_i)$ their empirical counterparts. They then estimate the density $f_Z(z)$ of the vector $Z = \{\Phi^{-1}(F_i(X_i))\}_{i=1,\ldots,p}$, in practice approximated by

@@

\begin{equation}
\mu_i(z_i) \stackrel{\textrm{def}}{=} 0 \,\, \textrm{ and } \sigma_i(z_i) \stackrel{\textrm{def}}{=} 1, \,\, i = 1,\ldots,p,
\label{eq:simp}
\end{equation}
so that the only parameter functions left to estimate are the pairwise local Gaussian correlations $R(z) = \{\rho_{ij}(z_i, z_j)\}_{i<j}$. We use the notation $Z$, $z_i$ and $z_j$ to signify that the estimation is performed on the (approximate) standard normal scale, or $z$-scale for short. We can then estimate the joint density $f_Z(z)$ of $Z$ as
\begin{equation}
\widehat f_Z(z) = \psi(z; \mu(z) = 0, \sigma(z) = 1, R = \widehat R(z)),
\label{eq:transformed-density}
\end{equation}
where $\mu(z) = \{\mu_i(z)\}$ and $\sigma(z) = \{\sigma_i(z)\}$ for $i=1,\ldots,p$, and then substitute $f_Z$ for $\widehat f_Z$ in the following relation obtained by \citet{otne:tjos:2017} in order to estimate the unknown density $f_X$:
\begin{equation}
f(x) = f_{Z}\big(\Phi^{-1}(F_1(x_1)), \ldots, \Phi^{-1}(F_p(x_p))\big) \times \prod_{i=1}^p \frac{f_i(x_i)}{\phi\big(\Phi^{-1}(F_i(x_i))\big)},
\label{eq:backtrans}
\end{equation}
where $\phi(\cdot)$ is the standard normal pdf. This estimator is implemented the \CRANpkg{lg}-package as demonstrated in Section \ref{chap:density}.

@@

\begin{align}
\mu^*(z) &= R_{12}(z)R_{22}(z)^{-1}z^{(2)},  \label{eq:cond-mu-trans}\\
\Sigma^*(z) &= R_{11}(z) - R_{12}(z)R_{22}(z)^{-1}R_{21}(z), \label{eq:cond-sig-trans}
\end{align}
where we, again, switch to $z$-notation in order to make it clear that these quantities are estimated on the standard normal $z$-scale. An estimator $\widehat f_{X^{(1)}|X^{(2)}}(\cdot|\cdot)$ of the conditional density $f_{X^{(1)}|X^{(2)}}(\cdot|\cdot)$ follows immediately from an expression corresponding to (\ref{eq:backtrans}), and the \CRANpkg{lg}-package provides functions for implementing this estimator in R. We describe the implementation of this functionality in Section \ref{chap:density}.

Finally, we refer to \citet{otne:tjos:2019} who take the local version of the conditional covariance matrix (\ref{eq:cond-sig}) (or (\ref{eq:cond-sig-trans}) in the transformed case) as a measure for conditional dependence, and thus as an instrument to test for conditional independence. Consider the stochastic vector $X = \left(X^{(1)}, X^{(2)}, X^{(3)}\right)$, where $X^{(1)}$ and $X^{(2)}$ are scalars and $X^{(3)}$ may be a vector. $X^{(1)}$ is conditionally independent from $X^{(2)}$ given $X^{(3)}$, written $X^{(1)} \perp X^{(2)} \,|\,X^{(3)}$, if the stochastic variables $X^{(1)}\,|\,X^{(3)}$ and $X^{(2)}\,|\,X^{(3)}$ are independent, or, equivalently, if the joint conditional density of $X^{(1)}$ and $X^{(2)}$ given $X^{(3)}$ can be written as the product

@@

The local Gaussian correlation may be used to perform a number of statistical analyses, as is evident from the preceding section. The practitioner must first, however, make three quite specific modeling choices; namely (i) to choose an estimation method, i.e. the level of simplification in multivariate applications, (ii) to determine whether the data should be transformed towards marginal standard normality before estimating the LGC, and (iii) to choose a set of bandwidths, or at least a method for calculating bandwidths. The architecture of the \pkg{lg}-packages requires the user to make these choices \emph{before} endeavoring further into specific applications by imposing a strict, two step procedure:

@@

In the following we assume that one has a \emph{data set} \code{x} loaded into the R workspace, which must be an $n\times p$ matrix (one column per variable, one row per observation), possibly including \code{NA}s which will be excluded from the analysis, or a data frame having the same dimensions. The fundamental syntax for creating an lg-object is \code{lg\_object <- lg\_main(x)}, and we will in this section explain how the modeling decisions (i)-(iii) can be encoded into the lg-object by using appropriate arguments in this function.

@@

Note that the full locally Gaussian fit for raw data is not available if the number of variables $p$ is greater than 2. The \code{lg\_main()}-function will check for this and print out an error message if $p>2$ and \code{est\_method = "5par"}.

@@

\textbf{3. A simplified locally Gaussian fit for marginally standard normal data.} This estimation method is applicable for marginally standard normal data, or data that have been transformed to approximate marginal standard normality by e.g. the transformation (\ref{eq:trans}). In that case, we fix the marginal expectation functions and standard deviation functions to the constant values $0$ and $1$ respectively, and estimate only the pairwise local Gaussian correlations as in (\ref{eq:transformed-density}). To use this estimation method, create the lg-object by running

@@

Note that the function call above will issue a warning if the option for transforming the data to marginal standard normality is not at the same time set to \code{TRUE}, see the next sub-section on data transformation for details.

\textbf{4. A full locally Gaussian fit for trivariate data.} If the number of variables $p$ is equal to $3$ \emph{and} we choose to transform the data to marginal standard normality (see the next sub-section), the transformed density $f_Z(\cdot)$ in (\ref{eq:transformed-density}) may be estimated by jointly estimating the three local correlations $\rho_{12}(z_1, z_2, z_3)$, $\rho_{13}(z_1, z_2, z_3)$ and $\rho_{23}(z_1, z_2, z_3)$. This estimation method was introduced recently by \citet{otne:tjos:2019} in order to increase power of their conditional independence test, but it can be used in any application described in this paper that consider trivariate data. To use this estimation method, create the lg-object by running

@@

We have in practice seen two automatic bandwidth selectors employed in the applications referred to in Section \ref{chap:background}: a
cross-validation procedure, that is fairly slow to compute, but accurate with respect to density estimation, and a plug-in bandwidth that is much quicker to calculate, but less accurate with respect to density estimation. We use the argument
\code{bw\_method} to the \code{lg\_main()}-function in order to choose between the two.

@@

$joint
x1 x2         bw1      bw2 convergence
1  1  2 0.2946889 0.331971           0
\end{example}
This is itself a list, containing the crucial elements \code{marginal} for the \(p\) marginal bandwidths, and \code{joint} that contains the
\(p(p-1)/2\) bandwidth matrices, one for each pair of variables (which in this bivariate example just \emph{one} variable pair, (\code{x1,\ x2})). The convergence flags stem from the built-in R functions \code{optim()} and \code{optimize()} that we use to obtain the minimizer of \(CV(\cdot)\), and 0 indicates successful convergence.

@@

In the sub-section above we present the three most important arguments to \code{lg\_main()}. Each of them allows the user to configure one of the three crucial modeling choices. Let us complete this treatment by covering some possibilities to make further adjustments to those choices.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The user may supply the bandwidths directly to \code{lg\_main()} by passing them to the \code{bw}-argument. They have to be in the   correct format though, which is a list containing the vector \code{\$marginal} if \code{est\_method\ =\ "5par\_marginals\_fixed"}, and always a data frame \code{\$joint} specifying all variable pairs in the \code{x1} and \code{x2} columns and the corresponding bandwidths
  in the \code{bw1} and \code{bw2} columns. The function \code{bw\_simple()} will assist in creating bandwidth objects.
\item
  If \code{bw\_method\ =\ "plugin"} the user may change the proportionality constant and exponent in the plugin formula for the joint and, if applicable, the marginal bandwidths. See Table \ref{tab:arguments-lg-main} for the necessary argument names.
\item
  If \code{bw\_method\ =\ "cv"}, the user may change the numerical tolerance in the optimization of \(CV(b)\). See Table \ref{tab:arguments-lg-main} for the necessary argument names.
\end{enumerate}

@@

We proceed in this section to demonstrate how to implement each of the tasks that we discussed in Section \ref{chap:background}. The general pattern is to pass the lg-object to one of the estimation or test functions provided in the \CRANpkg{lg}-package. We will look at some financial data in the examples: the monthly returns on the S\&P500, FTSE100, DAX30 and TOPIX stock indices from January 1985 to March 2018. The data is available as an online appendix to this article.
